---
title: "Basic usage of inedemogR"
output: html_document
---

```{r setup, include=FALSE}
# Chunks are not evaluated: examples hit the live INE API and require the
# inedemogR package, which is not installed when this site is built.
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, eval = FALSE)
```

```{r}
library(inedemogR)
```

`inedemogR` wraps the Spanish National Statistics Institute (INE) "fenómenos
demográficos" tables (population, births, deaths) behind a small set of tidy
retrieval functions, a mortality/fertility analysis pipeline, and `ggplot2`
plotting helpers. Data is fetched live via `ineapir`; a local cache under
`tools::R_user_dir("inedemogR", "cache")` avoids re-downloading unchanged tables.

## 1. Quick multi-geography totals

`list_ine_indicators()` shows the headline indicators, and `get_ine_demog()`
returns them tidied to one row per geography × year.

```{r}
list_ine_indicators()

# Resident population per municipality, 2023
pop_totals <- get_ine_demog(indicator = "population_total", year = 2023)

# Births and deaths together, province level, one call
vitals <- get_ine_demog(
  indicator = c("births_total", "deaths_total"),
  year = 2023
)
```

Pass `geometry = TRUE` to get an `sf` object with `mapSpain` boundaries joined,
ready to plot directly.

```{r}
pop_sf <- get_ine_demog("population_total", year = 2023, geometry = TRUE)
```

## 2. Age/sex-disaggregated series

The System B retrieval functions pull the longer age-specific province series
that feed the analysis pipeline. Each returns a list with a `$data` tibble plus
quality-control tables.

```{r}
pop        <- get_ine_population(n_periods = 10)
births     <- get_ine_births()
births_age <- get_ine_births_by_age()
deaths     <- get_ine_deaths()
```

## 3. Summary demographic indicators

```{r}
adr <- age_dependency_ratio(pop$data)
ai  <- aging_index(pop$data)

cbr <- crude_birth_rate(births$data, pop$data)
cdr <- crude_death_rate(deaths$data_provinces, pop$data)
rni <- rate_of_natural_increase(cbr, cdr)
```

## 4. Fertility schedule

```{r}
asfr <- age_specific_fertility_rate(births_age$data, pop$data, age_min = 15, age_max = 49)
tfr  <- total_fertility_rate(asfr)
mac  <- mean_age_at_childbearing(asfr)
```

## 5. Mortality pipeline and life tables

```{r}
exposure <- compute_exposure(pop$data, deaths$data_provinces)
rates    <- compute_death_rates(deaths$data_provinces, exposure$data)

lt <- build_life_tables(rates$mx_1x1)          # 1x1 period life tables, all provinces/years
le <- life_expectancy_summary(lt$fltper, sex = "female")
```

## 6. Visualization

The plotting helpers return `ggplot` objects built from the retrieval/pipeline
output:

```{r}
# Mirrored age/sex population pyramid for one province
plot_population_pyramid(pop$data, year = max(pop$data$year), region = "A Coruna")

# Trend of an indicator over time
plot_demog_trend(ai, "aging_index", region = "A Coruna")

# Age × year mortality surface with cohort diagonals
plot_lexis_diagram(rates$mx_1x1, province = "Madrid")

# Choropleth of any value column
map_indicator(pop_totals, "population_total", geo_level = "province")
```

The figures below are rendered from small **illustrative** data frames (not a
live INE download) shaped like the real output, to show what the helpers produce.

```{r pyramid, eval=TRUE, echo=TRUE, fig.height=5, fig.cap="Population pyramid (illustrative data)"}
library(ggplot2)

set.seed(1)
ages <- 0:99
decay <- exp(-((ages - 42)^2) / 2200)
pyr <- data.frame(
  age  = rep(ages, 2),
  sex  = rep(c("female", "male"), each = length(ages)),
  count = c(round(3800 * decay * runif(length(ages), 0.9, 1.1)),
            round(3600 * decay * runif(length(ages), 0.9, 1.1)))
)
pyr$signed <- ifelse(pyr$sex == "male", -pyr$count, pyr$count)

ggplot(pyr, aes(x = age, y = signed, fill = sex)) +
  geom_col(width = 1) +
  coord_flip() +
  scale_y_continuous(labels = function(x) format(abs(x), big.mark = ",")) +
  scale_fill_manual(values = c(female = "#c0392b", male = "#2c3e50")) +
  labs(x = "Age", y = "Population", fill = NULL,
       title = "A Coruña, 2023 (illustrative)") +
  theme_minimal()
```

```{r trend, eval=TRUE, echo=TRUE, fig.height=4, fig.cap="Aging index trend (illustrative data)"}
trend <- data.frame(
  year = 2002:2023,
  aging_index = seq(95, 210, length.out = 22) + rnorm(22, 0, 4)
)

ggplot(trend, aes(year, aging_index)) +
  geom_line(colour = "#2c3e50", linewidth = 1) +
  geom_point(colour = "#2c3e50") +
  labs(x = NULL, y = "Aging index (65+ per 100 under-15)",
       title = "A Coruña (illustrative)") +
  theme_minimal()
```

```{r asfr, eval=TRUE, echo=TRUE, fig.height=4, fig.cap="Age-specific fertility rate (illustrative data)"}
asfr_demo <- data.frame(
  age = 15:49,
  rate = dnorm(15:49, mean = 32, sd = 5.5) * 2.1
)

ggplot(asfr_demo, aes(age, rate)) +
  geom_area(fill = "#c0392b", alpha = 0.25) +
  geom_line(colour = "#c0392b", linewidth = 1) +
  labs(x = "Age of mother", y = "Births per woman",
       title = "ASFR schedule, TFR ≈ 2.1 (illustrative)") +
  theme_minimal()
```

### An indicator on the map of Spain

`get_ine_demog(..., geometry = TRUE)` and `map_indicator()` return province-level
`sf`/choropleth output. Here the geometry comes straight from `mapSpain` (as it
does inside `inedemogR`), with a **simulated** aging-index column joined on.

```{r map, eval=TRUE, echo=TRUE, fig.height=6, fig.cap="Aging index by province (simulated values, real mapSpain geometry)"}
library(mapSpain)
library(sf)

prov <- esp_get_prov(moveCAN = TRUE)          # 52 provinces, Canaries shifted
can_box <- esp_get_can_box(moveCAN = TRUE)

set.seed(42)
prov$aging_index <- round(runif(nrow(prov), 60, 260))

ggplot(prov) +
  geom_sf(aes(fill = aging_index), colour = "white", linewidth = 0.1) +
  geom_sf(data = can_box, colour = "grey40") +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "Aging index") +
  labs(title = "Aging index by province (simulated)") +
  theme_void()
```

## 7. Exporting

`download_ine_data()` runs whichever pipeline stages you ask for (resolving
their dependencies) and writes the results as CSV and/or HMD-format `.txt`.

```{r}
download_ine_data("ine_data", stages = c("births", "deaths"), format = "csv")
```

See the full `inedemogR` tutorial (`vignettes/inedemogR-tutorial.pdf` in the
[package repository](https://github.com/jrcarob/inedemogR_package)) for the
methodology behind each function.
